Sparse spectral and $p$-finite element methods for partial differential equations on spherical caps

### 2020-08-06 papers/spherical-caps/spherical-caps.tex
Further edits to paper. Need to update images and ensure prose is correct.

% Tangent space
\newcommand{\tangentspace}{{\Omega_T}}
\newcommand{\phivec}{\hat{\bm{\phi}}}
\newcommand{\thetavec}{\hat{\bm{\theta}}}
\newcommand{\bigtsop}{\mathbb{T}}
\newcommand{\bigtsopt}{\tilde{\bigtsop}}
\newcommand{\bigtsoptn}{\bigtsopt_n}
\newcommand{\bigtsopN}{\bigtsop_N}
\newcommand{\bigtsopNk}{\bigtsop_{N,k}}
\newcommand{\tsopi}{\bm{\Phi}}
\newcommand{\tsopii}{\bm{\Psi}}
\newcommand{\tsopiN}{\tsopi_N}
\newcommand{\tsopiiN}{\tsopii_N}
\newcommand{\tsopinki}{\tsopi_{n,k,i}}
\newcommand{\tsopiinki}{\tsopii_{n,k,i}}
\newcommand{\tsopinkit}{\tsopi_{n,k,i}^\top}
\newcommand{\tsopiinkit}{\tsopii_{n,k,i}\top}
\newcommand{\tsopit}{\tsopi^\top}
\newcommand{\tsopiit}{\tsopii^\top}
\newcommand{\rvec}{\hat{\bm{r}}}
\newcommand{\jacobimattangent}{{}^T\!J}
\newcommand{\jacobimattangentx}{\jacobimattangent_x}
\newcommand{\jacobimattangenty}{\jacobimattangent_y}
\newcommand{\jacobimattangentz}{\jacobimattangent_z}
\newcommand{\alphao}{\alpha^{(0)}}
\newcommand{\alphaonkj}{\alphao_{n,k,j}}
\newcommand{\betao}{\beta^{(0)}}
\newcommand{\betaonkj}{\betao_{n,k,j}}
\newcommand{\gammao}{\gamma^{(0)}}
\newcommand{\gammaonkj}{\gammao_{n,k,j}}

@@

\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma} 
\newtheorem{theorem}{Theorem} 
\newtheorem{definition}{Definition}
\newtheorem{condition}{Condition}
\newtheorem{algorithm}{Algorithm}
\newtheorem{corollary}{Corollary}

@@

\section{Introduction}

We can use the similar idea of projecting up from the 2D orthogonal polynomials on the half disk to 3D OPs on the sphere. Define the disk-slice, denoted by $\omega$, by
\begin{align}
	\omega := \{(x,y) \in \R^2 \quad | \quad \alpha \le x \le \beta, \: \gamma \rho(x) \le y \le \delta \rho(x)\}.
\end{align}
where:
\begin{align*}
	\begin{cases}
		(\alpha,\beta) &:= (\alpha,1), \quad \alpha \ge 0 \\
		(\gamma,\delta) &:= (-1,1) \\
		\rho(x) &:= (1-x^2)^{\half}, \quad x \in (\alpha,\beta) \\
	\end{cases}
\end{align*}

Further, define the spherical cap, denoted by $\Omega$, by
\begin{align}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad (x,y) \in \omega, \: x^2 + y^2 + z^2 = 1\}.
\end{align}
We can define the 3D OP family on $\Omega$ as the set of polynomials $\bigQab(x,y,z)$ where
\begin{align}
	\bigQab_n(x,y,z) := \begin{pmatrix}
		\bighdopab_n(x,y) \\
		z \: \bighdop^{(a,b+1)}_{n-1}(x,y)
	\end{pmatrix} \in \R^{2n+1}, 
	\quad \quad 
	\bigQab := \begin{pmatrix}
		\bigQab_0 \\
		\bigQab_1 \\
		\vdots \\
		\bigQab_N
	\end{pmatrix}.
\end{align}
The relations for multiplication by $x, y$ are then trivial. For multiplication by $z$, note that $z^2 = W^{(0,1)}(x,y)$ for $(x,y,z) \in \Omega$ and consider that (as the 2D OPs for any parameters form a 2D basis of $\omega$):
\begin{align}
	z \: \hdopnkab(x,y) &= \sum_{m,j} c_{n,k,m,j}^{(a,b)\to(a,b+1)} \: z \: \hdop^{(a,b+1)}_{m,j}(x,y), \\
	z^2 \: \hdopnk^{(a,b+1)}(x,y) &= \sum_{m,j} d_{n,k,m,j}^{(a,b+1)\to(a,b)} \: \hdopmjab(x,y).
\end{align}
The coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ and $d_{n,k,m,j}^{(a,b+1)\to(a,b)}$ are given by
\begin{align}
	c_{n,k,m,j}^{(a,b)\to(a,b+1)} &= \frac{\ip< \hdopnkab, \: \hdop^{(a,b+1)}_{m,j} >_{W^{(a,b+1)}}}{\norm{\hdop^{(a,b+1)}_{m,j}}^2_{W^{(a,b+1)}}} \\
	d_{n,k,m,j}^{(a,b+1)\to(a,b)} &= \frac{\ip< W^{(0,1)} \hdopnk^{(a,b+1)}, \: \hdopmjab >_\Wab}{\norm{\hdopmjab}^2_\Wab} \nonumber \\
        	&= \frac{\ip< \hdopnk^{(a,b+1)}, \: \hdopmjab >_{W^{(a,b+1)}}}{\norm{\hdopmjab}^2_\Wab} \nonumber \\
        	&= \frac{\norm{\hdopnk^{(a,b+1)}}^2_{W^{(a,b+1)}}}{\norm{\hdopmjab}^2_\Wab} \: c_{m,j,n,k}^{(a,b)\to(a,b+1)}.
\end{align}
The coefficients $c_{n,k,m,j}^{(a,b)\to(a,b+1)}$ are just the entries to the parameter transformation operator matrix $T^{(a,b)\to(a,b+1)}$, defined by
\begin{align}
	\bighdop^{(a,b)}(x,y) &= \Big(T^{(a,b)\to(a,b+1)} \Big)^T \: \bighdop^{(a,b+1)}(x,y),
\end{align}
which can be shown to be sparse with banded-block-banded structure, meaning the resulting operator matrix for the Jacobi operator for multiplication by $z$ is sparse.

@@

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: ||(x,y)|| = \gamma \rho(z)\}
\end{align*}
and consider the case where $\alpha \in (0,1)$ and where
\begin{align*}
	 \rho(z) &:= \sqrt{1-z^2} \\
	 \beta &:= 1 \\
	 \gamma &:= 1.
\end{align*}
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < 1, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
Of course, there is a simple extension to a spherical band by taking $\beta \in (\alpha,1)$, however for simplicity we focus on the spherical cap.

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the disk \cite{vasil2016tensor} and triangle \cite{olver2018recurrence,olver2019triangle}.  On the disk-slice the family of weights we consider are of the form
$$
\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \qfor \alpha \leq x \leq \beta,\quad -\rho(x) \leq y \leq \rho(x).
$$
The corresponding OPs denoted $\hdopnkabc(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkabc(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms''   includes degree $n$ polynomials of the form $x^{n-j} y^j$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator using different choices of the parameters $a$, $b$ and $c$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle and \cite{vasil2016tensor} for equations on the disk results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside careful use of the Christoffel--Darboux formula \cite[18.2.2]{DLMF} and quadrature rules to determine the entries. In particular, by exploiting the connection with one-dimensional orthogonal polynomials we can construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

Here is an overview of the paper:

\noindent \secref{OPs}: We present our procedure to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the disk-slice domain, by combining 1D OPs on the interval, to form 2D OPs on the disk.

\noindent\secref{PDOs}: We demonstrate that these families will lead to sparse operators, including Jacobi operators representing multiplication by $x$ and $y$, and partial differential operators. We present a method involving use of the Christoffel--Darboux formula \cite[18.2.2]{DLMF} to obtain the recurrence coefficients for the non-classical 1D OPs, allowing us to exactly use the 1D quadrature rules we present to calculate the non-zero entries to the sparse operators.

\noindent\secref{Computation}: We discuss computational issues, in particular, how to realise the results of the preceding sections in practice. We present a method for explicitly deriving the recurrence coefficients of the non-classical 1D OPs we detail in \secref{OPs}. We derive a quadrature rule on the disk-slice that can be used to expand a function in the OP basis up to a given order. Further, we implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm.

\noindent\secref{Examples}: We demonstrate the proposed technique for solving Poisson, Helmholtz, and Biharmonic equations on the disk-slice.

\noindent\appref{PFEM}: We use the procedure to construct sparse $p$-finite element methods. This lays the groundwork for a future $hp$-finite element method in a disk, where the elements capture the circular geometry precisely.

\noindent\appref{HalfDisk}: We discuss extension to the special case of end-disk-slices (e.g., half disks).

\noindent\appref{Trapezium}: We discuss extension to trapezia.

@@

We can construct the 3D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the interval \([\alpha,\beta]\), and from the 2D circular harmonics on the unit circle.

\begin{definition}
Define $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ as the 2D OPs on $(x,y)\in\R$ s.t. $x^2 + y^2 = 1$ (known as the circular harmonics on the unit circle); that is, if we define $\theta$ by $x = \cos(\theta)$, $y = \sin(\theta)$ for $\xvec := (x, y)$ where $||\xvec|| = 1$,
\begin{align}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) \equiv \ch_{0,0}(\theta) &:= \frac{\sqrt{2}}{2} =: \ch_{0} \nonumber \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \equiv \ch_{k,0}(\theta) &:= \cos(k \theta), \quad k = 1,2,3,\dots\label{eqn:circularharmonics} \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) \equiv \ch_{k,0}(\theta)& := \sin(k \theta), \quad k = 1,2,3,\dots \nonumber
\end{align}
orthonormal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\sqrt{\pi}} \: \int_{||\xvec|| = 1} p(\xvec) \: q(\xvec) \: \D \sigma(\xvec) = \frac{1}{\sqrt{\pi}} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta. \label{eqn:ipcircularharmonics}
\end{align}
\end{definition}

@@

For the spherical cap, we formalise Proposition \ref{prop:construction} in the following definitions. First we introduce notation for two families of univariate OPs.
\begin{definition}\label{def:OPconstruction}
Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, \beta)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{b} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^\beta \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b,c)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}).
\end{definition}

\begin{definition}\label{def:constuction}
Define the one-parameter 3D orthogonal polynomials via:
\begin{align}
	\scopnkia(x,y,z) := \genjacnmk^{(a,2k)}(x) \: \rho(x)^k \: \chki\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
\end{definition}

$\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z.
\end{align}

We note that the weight $w_R(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}), and has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}.

@@

\begin{lemma}\label{lemma:Yrecurrence} 
The following identities hold for $k = 2,3,\dots$:
\begin{align*}
	&1) \quad \int_0^{2\pi} \ch_0 \: \chjh(\theta) \: \cos(\theta) \: \D \theta = \ch_0 \: \pi \: \delta_{0,h} \: \delta_{1,j} \\
	&2) \quad \int_0^{2\pi} \ch_0 \: \chjh(\theta) \: \sin(\theta) \: \D \theta = \ch_0 \: \pi \: \delta_{1,h} \: \delta_{1,j} \\
	&3) \quad \int_0^{2\pi} \ch_{1,i}(\theta) \: \chjh(\theta) \: \cos(\theta) \: \D \theta = \pi \: \delta_{i,h} \: (\ch_0 \: \delta_{0,j} + \half \delta_{2,j}) \\
	&4) \quad \int_0^{2\pi} \ch_{1,i}(\theta) \: \chjh(\theta) \: \sin(\theta) \: \D \theta = \pi \:  \delta_{|i-1|,h} \: ((-1)^{i+1} \: \ch_0 \: \delta_{0,j} + (-1)^i \: \half \: \delta_{2,j}) \\
	&5) \quad \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \cos(\theta) \: \D \theta = \half \: \pi \: \delta_{i,h} \: (\delta_{k-1,j} + \delta_{k+1,j}) \\
	&6) \quad \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \sin(\theta) \: \D \theta = \half \: \pi \: \delta_{|i-1|,h} \: ((-1)^{i+1} \: \delta_{k-1,j} + (-1)^i \: \delta_{k+1,j}).
\end{align*}
\end{lemma}
\begin{proof}
Each follows from the definitions of $\ch_{k,i}$ and $\ch_0$, as well as the relationships:
\begin{align*}
	2 \: \cos(k \theta) \cos(\theta) &= \cos((k-1)\theta) + \cos((k+1)\theta) \\
	2 \: \sin(k \theta) \cos(\theta) &= \sin((k-1)\theta) + \sin((k+1)\theta) \\
	2 \: \cos(k \theta) \sin(\theta) &= - \sin((k-1)\theta) + \sin((k+1)\theta) \\
	2 \: \sin(k \theta) \sin(\theta) &= \cos((k-1)\theta) - \cos((k+1)\theta).
\end{align*}
\end{proof}

>>>>

% Tangent space
\newcommand{\tangentspace}{{\Omega_T}}
\newcommand{\phivec}{\hat{\bm{\phi}}}
\newcommand{\thetavec}{\hat{\bm{\theta}}}
\newcommand{\bigtsop}{\mathbb{T}}
\newcommand{\bigtsopt}{\tilde{\bigtsop}}
\newcommand{\bigtsoptn}{\bigtsopt_n}
\newcommand{\bigtsopN}{\bigtsop_N}
\newcommand{\bigtsopNk}{\bigtsop_{N,k}}
\newcommand{\tsopi}{\bm{\Phi}}
\newcommand{\tsopii}{\bm{\Psi}}
\newcommand{\tsopiN}{\tsopi_N}
\newcommand{\tsopiiN}{\tsopii_N}
\newcommand{\tsopinki}{\tsopi_{n,k,i}}
\newcommand{\tsopiinki}{\tsopii_{n,k,i}}
\newcommand{\tsopinkit}{\tsopi_{n,k,i}^\top}
\newcommand{\tsopiinkit}{\tsopii_{n,k,i}\top}
\newcommand{\tsopit}{\tsopi^\top}
\newcommand{\tsopiit}{\tsopii^\top}
\newcommand{\rvec}{\hat{\bm{r}}}
\newcommand{\jacobimattangent}{{}^T\!J}
\newcommand{\jacobimattangentx}{\jacobimattangent_x}
\newcommand{\jacobimattangenty}{\jacobimattangent_y}
\newcommand{\jacobimattangentz}{\jacobimattangent_z}
\newcommand{\alphao}{\alpha^{(0)}}
\newcommand{\alphaonkj}{\alphao_{n,k,j}}
\newcommand{\betao}{\beta^{(0)}}
\newcommand{\betaonkj}{\betao_{n,k,j}}
\newcommand{\gammao}{\gamma^{(0)}}
\newcommand{\gammaonkj}{\gammao_{n,k,j}}
\newcommand{\uvec}{\bm{u}}
\newcommand{\hvec}{\bm{h}}
\newcommand{\uvecc}{\uvec^c}
\newcommand{\hvecc}{\hvec^c}
\newcommand{\uvecn}{\uvec_n}
\newcommand{\hvecn}{\hvec_n}
\newcommand{\uvecnpi}{\uvec_{n+1}}
\newcommand{\hvecnpi}{\hvec_{n+1}}
\newcommand{\uvecnc}{\uvecc_n}
\newcommand{\hvecnc}{\hvecc_n}
\newcommand{\uvecnpic}{\uvecnpi^c}
\newcommand{\hvecnpic}{\hvecnpi^c}
\newcommand{\href}{\mathcal{H}}
\newcommand{\divergence}{\nabla \cdot}
\newcommand{\grad}{\nabla}

@@

\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma} 
\newtheorem{theorem}{Theorem} 
\newtheorem{definition}{Definition}
\newtheorem{condition}{Condition}
\newtheorem{algorithm}{Algorithm}
\newtheorem{corollary}{Corollary}
\newtheorem*{remark}{Remark}

@@

@@

This paper develops sparse spectral methods for solving linear partial differential equations on spherical subdomains such as spherical bands and spherical caps. More precisely, we consider the solution of partial differential equations on the domain $\Omega$ given by
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < \beta, \: ||(x,y)|| = \gamma \rho(z)\}
\end{align*}
and consider the case where $\alpha \in (0,1)$ and where
\begin{align*}
	 \rho(z) &:= \sqrt{1-z^2} \\
	 \beta &:= 1 \\
	 \gamma &:= 1.
\end{align*}
This framework then yields that $\Omega$ is a spherical cap (the region of the surface of a sphere where the $z$-coordinate ranges from $\alpha$ to $1$), i.e.
\begin{align*}
	\Omega := \{(x,y,z) \in \R^3 \quad | \quad \alpha < z < 1, \: x^2 + y^2 + z^2 = 1\}
\end{align*}
Of course, there is a simple extension to a spherical band by taking $\beta \in (\alpha,1)$. However for simplicity we focus on the spherical cap in this paper.

As in the work in two dimensions on the show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that are defined in such a way that mirrors the approach on quadratic curves \cite{olver2020orthogonal}. On the spherical cap, the family of weights we consider are of the form
\begin{align*}
	\Wa(x,y,z) := (z - \alpha)^a, \qfor (x,y,z) \in \Omega,
\end{align*}
noting that $\Wa(x,y,z) = 0$ for $(x,y,z) \in \partial \; \Omega$. The corresponding OPs denoted $\scopnkia(x,y,z)$, where $n$ denotes the polynomial degree, $0 \le k \le n$ and $i \in \{0, \min(1,k)\}$. We define these to be orthogonalised lexicographically, that is,
\begin{align*}
	\scopnkia(x,y,z) = C_{n,k,i} \: x^{k - \tilde k_i} \: y^{\tilde k_i} \: z^{n-k} + (\hbox{lower order terms})
\end{align*} \bstodo{Ordering correct?}
where $C_{n,k,i} \neq 0$ and ``lower order terms''   includes degree $n$ polynomials of the form $x^{j - \tilde j_i} \: y^{\tilde j_i} \: z^{n-j}$ where $j < k$. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator using different choices of the parameter $a$. In this work, as for the disk-slice case in 2D \cite{snowball2019sparse}, we use a more general integration-by-parts argument to deduce the sparsity structure. The 3D OPs defined here involve the same non-classical (in fact, semi-classical) 1D OPs as those outlined for the disk-slice, and so methods for calculating these 1D OP recurrence coefficients and integrals has already been outlined \cite{snowball2019sparse}. In particular, by exploiting the connection with one-dimensional orthogonal polynomials we can construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}. \bstodo{reword to avoid plagiarising self}

An overview of the paper is as follows:

\noindent \secref{OPs}: We present our definition of a (one-parameter) family of 3D orthogonal polynomials (OPs) on the spherical cap domain $\Omega$, by combining 1D OPs on the interval $(\alpha, 1)$ with Chebyshev polynomials, to form 3D OPs on the spherical cap surface. We show that these families will lead to sparse Jacobi operators for multiplication by $x, y, z$ and demonstrate how to obtain the 3D OPs.

\noindent\secref{PDOs}: We define useful partial differential operators and further show that these will also be sparse. We can exactly calculate the non-zero entries of these sparse operators using the quadrature rule associated with the non-classical 1D OPs.

\noindent\secref{Computation}: We derive a quadrature rule on the spherical cap that can be used to expand a function in the OP basis up to a given order $N$, and demonstrate how to evaluate a function using the Clenshaw algorithm using the coefficients of its expansion.

\noindent\secref{Examples}: We demonstrate the proposed technique for solving differential equations on the spherical cap such as the Poisson equation, variable coefficient Helmholtz equation, and Biharmonic equation.

\noindent\appref{PFEM}: We outline how our framework can allow us to construct sparse $hp$-finite element method on the sphere, using spherical band and cap elements that would capture spherical geometry precisely. We lay the groundwork by simply presenting a sparse $p$-finite element method using a single spherical cap element.

@@

We can construct the 3D orthogonal polynomials on $\Omega$ from 1D orthogonal polynomials on the interval \([\alpha,\beta]\), and from Chebyshev polynomials.

\begin{definition}
Define $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ as the 2D OPs on $(x,y)\in\R$ s.t. $x^2 + y^2 = 1$; that is, if we define $\theta$ by $x = \cos(\theta)$, $y = \sin(\theta)$ for $\xvec := (x, y)$ where $||\xvec|| = 1$,
\begin{align}
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) \equiv \ch_{0,0}(\theta) &:= \frac{\sqrt{2}}{2} =: \ch_{0} \nonumber \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) \equiv \ch_{k,0}(\theta) &:= \cos(k \theta), \quad k = 1,2,3,\dots\label{eqn:circularharmonics} \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) \equiv \ch_{k,0}(\theta)& := \sin(k \theta), \quad k = 1,2,3,\dots \nonumber
\end{align}
orthonormal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_{||\xvec|| = 1} p(\xvec) \: q(\xvec) \: \D \sigma(\xvec) = \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta. \label{eqn:ipcircularharmonics}
\end{align}
\end{definition}
Of course, these are simply Chebyshev polynomials $\ch_{k,0}(x, y) = T_k(x)$ and $\ch_{k,1}(x, y) = y \: U_{k-1}(x)$ for $k > 0$, where we define $\ch_0$ so as to ensure orthonormality.

@@

For the spherical cap, we can use Proposition \ref{prop:construction} to create our one-parameter family of OPs. We first introduce notation for our family of non-classical univariate OPs that will be used as the $r_n$ polynomials above.
\begin{definition}[\cite{snowball2019sparse}]\label{def:OPconstruction}
Let $\genjacw^{(a,b)}(x)$ be a weight function on the interval $(\alpha, \beta)$ given by:
\begin{align*}
	\genjacw^{(a,b)}(x) &:= (x - \alpha)^{b} \: \rho(x)^{b}
\end{align*}
and define the associated inner product by:
\begin{align}
	\ip< p, \: q >_{\genjacw^{(a,b)}} &:= \frac{1}{\normgenjac^{(a,b)}} \: \int_\alpha^\beta p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipgenjac}
\end{align}
where
\begin{align}
	\normgenjac^{(a,b)} := \int_\alpha^\beta \: \genjacw^{(a,b)}(x) \: \D x \label{eqn:ipnormalisation}
\end{align}
is a normalising constant.
Denote the two-parameter family of orthonormal polynomials on $[\alpha,\beta]$ by $\{\genjac_n^{(a,b)}\}$, orthonormal with respect to the inner product defined in (\ref{eqn:ipgenjac}).
\end{definition}

We can now define the 3D OPs for the spherical cap.
\begin{definition}\label{def:constuction}
Define the one-parameter 3D orthogonal polynomials via:
\begin{align}
	\scopnkia(x,y,z) := \genjacnmk^{(a,2k)}(x) \: \rho(x)^k \: \chki\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
\end{definition}

By construction, $\{\scopnkia\}$ are orthogonal with respect to the inner product
\begin{align}
	\ip< p, \: q >_{\scopa} &:= \int_\Omega p(\xvec, z) \: q(\xvec, z) \: \genjacw^{(a,0)}(z) \: \D \sigma(\xvec, z) \\
	&= \int_\alpha^1 \int_0^{2\pi} p(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: q(\rho(z) \cos(\theta), \rho(z) \sin(\theta), z) \: \D \theta \: \genjacw^{(a,0)}(z) \: \D z,
\end{align}
with
\begin{align}
	\norm{\scopnkia}_{\scopa}^2 &:= \ip< \scopnkia, \: \scopnkia >_{\scopa} \nonumber \\
	&= \pi \: \normgenjac^{(a, 2k)} \label{eqn:normscop}
\end{align}
We note that the weight $\genjacw^{(a,b)}(z)$ has been used in the construction of 2D orthogonal polynomials on disk-slices and trapeziums \cite{snowball2019sparse}, where a method for obtaining recurrence coefficients and evaluating integrals was established (the weight is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}).

@@

\begin{lemma}\label{lemma:Yrecurrence} 
The following identities hold for $k = 2,3,\dots$, $j = 0,1,\dots$ and $i, h \in \{0,1\}$:
\begin{align*}
	&1) \quad \int_0^{2\pi} \ch_0 \: \chjh(\theta) \: \cos(\theta) \: \D \theta = \ch_0 \: \pi \: \delta_{0,h} \: \delta_{1,j} \\
	&2) \quad \int_0^{2\pi} \ch_0 \: \chjh(\theta) \: \sin(\theta) \: \D \theta = \ch_0 \: \pi \: \delta_{1,h} \: \delta_{1,j} \\
	&3) \quad \int_0^{2\pi} \ch_{1,i}(\theta) \: \chjh(\theta) \: \cos(\theta) \: \D \theta = \pi \: \delta_{i,h} \: (\ch_0 \: \delta_{0,j} + \half \delta_{2,j}) \\
	&4) \quad \int_0^{2\pi} \ch_{1,i}(\theta) \: \chjh(\theta) \: \sin(\theta) \: \D \theta = \pi \:  \delta_{|i-1|,h} \: ((-1)^{i+1} \: \ch_0 \: \delta_{0,j} + (-1)^i \: \half \: \delta_{2,j}) \\
	&5) \quad \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \cos(\theta) \: \D \theta = \half \: \pi \: \delta_{i,h} \: (\delta_{k-1,j} + \delta_{k+1,j}) \\
	&6) \quad \int_0^{2\pi} \chki(\theta) \: \chjh(\theta) \: \sin(\theta) \: \D \theta = \half \: \pi \: \delta_{|i-1|,h} \: ((-1)^{i+1} \: \delta_{k-1,j} + (-1)^i \: \delta_{k+1,j}).
\end{align*}
\end{lemma}
\begin{proof}
Each follows from the definitions of $\ch_{k,i}$ and $\ch_0$, as well as the relationships:
\begin{align*}
	2 \cos(k \theta) \cos(\theta) &= \cos((k-1)\theta) + \cos((k+1)\theta) \\
	2 \sin(k \theta) \cos(\theta) &= \sin((k-1)\theta) + \sin((k+1)\theta) \\
	2 \cos(k \theta) \sin(\theta) &= - \sin((k-1)\theta) + \sin((k+1)\theta) \\
	2 \sin(k \theta) \sin(\theta) &= \cos((k-1)\theta) - \cos((k+1)\theta).
\end{align*}
\end{proof}
